Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-20 Paper/app.tex
..

For the high-dimensional case, we aim to investigate the impact of the uncertain
pre-exponents ($A_i$'s) as well as the activation energies ($E_{a,i}$'s) on ignition 
delay during the H$_2$/O$_2$ reaction. The $A_i$'s were considered to be uniformly
distributed in the interval, $[0.9A_i^\ast, 1.1A_i^\ast]$. The $E_{a,i}$'s for all
reactions except $\mathcal{R}_6$ -- $\mathcal{R}_9$ and $\mathcal{R}_{13}$ (due to 
zero nominal values for $E_a$)
were considered to be uncertain and uniformly distributed in the interval: 
$[0.99E_{a,i}^\ast, 1.01E_{a,i}^\ast]$. The nominal values, $A_i^\ast$ and $E_{a,i}^\ast$
corresponding to the different reaction rates are provided in~\cite{Yetter:1991}.

@@

In Figure~\ref{fig:hd} (left), we illustrate the resulting eigenvalue spectrum. The second eigenvalue is
found to be roughly two orders of magnitude smaller than the first, indicating that the active subspace
is 1-dimensional. This is further confirmed by the SSP plots in Figure~\ref{fig:hd} (right). Specifically
in the case of gradient-based approach, the SSP exhibits a linear variation that is reasonably
captured by its linear fit, $\tilde{G}$. Whereas, the SSP based on the gradient-free approach
is relatively more scattered. Consequently, the linear fit exhibits a discrepancy with the corresponding
fit for the gradient-based SSP. 
%
\begin{figure}[htbp]
 \begin{center}
   \includegraphics[width=0.45\textwidth]{./Figures/eig_33D}
   \includegraphics[width=0.45\textwidth]{./Figures/ssp_33D}
\caption{Left: Eigenvalue spectrum of $\hat{\mat{C}}$. Right: SSP of the computed active subspace, and the
corresponding 1-D linear fit.} 
\label{fig:hd}
\end{center}
\end{figure}
%
We plot the normalized activity scores as well as the total Sobol' indices for the 33 uncertain 
rate-controlling parameters in Figure~\ref{fig:as_33D}. Note that the normalized activity scores
as well as the Sobol' indices
were evaluated using respective 1-dimensional surrogates for the two approaches.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/as_33D_new}
\caption{A bar-graph illustrating individual activity scores for the uncertain $A_i$' (left) and $E_{a,i}$'s (right).}
\label{fig:as_33D}
\end{center}
\end{figure}
%

Several useful inferences can be drawn from the above plot. Firstly, the total Sobol indices are found to be
consistent
with the normalized activity scores evaluated uing the gradient-based approach. Whereas, the gradient-free
approach yields consistent results for the $A_i$'s, it does not capture the sensitivity associated with
$E_{a,15}$. While this observation bolsters our confidence in the gradient-based approach, it clearly
underscores the shortcoming of the gradient-free approach. Numerical errors are incurred when approximating
the model gradients using the local linear approximation in this case. In~\ref{sub:verify}, we further assess the
suitability of the gradient-free approach in quantifying the uncertainty in the ignition delay. 
Secondly, our findings indicate that   
the ignition delay is predominantly sensitive towards $A_1$, $A_9$, $E_{a,1}$, and $E_{a,15}$ and 
moderately sensitive towards $A_{15}$ and $A_{17}$. Sensitivity towards the remaining uncertain inputs is
found to be low or negligible. Note that this observation can also be exploited for reducing the 
dimensionality of the problem from 33 inputs to 4-6 inputs depending upon the required level of accuracy.
The sensitivity-driven dimension reduction for uncertainty quantification has been
discussed in our earlier effort~\cite{Vohra:2018}. In this work, however, we focus on dimension reduction using the active 
subspace approach which is shown to yield even greater scope for dimension reduction i.e. from 33 to 1. Furthermore, we have
shown that the dominant eigenspace can be used to approximate global sensitivity measures with no additional effort.
The accuracy of the 1-dimensional surrogate, $\tilde{G}$, obtained using the two approaches is assessed in the following section.

@@

%\subsection{Surrogate-induced risk analysis}

>>>>

For the high-dimensional case, we aim to investigate the impact of the uncertain
pre-exponents ($A_i$'s) as well as the activation energies ($E_{a,i}$'s) on ignition 
delay during the H$_2$/O$_2$ reaction.  Prior distributions for $A_i$'s were considered
to be the same as provided earlier in section~\ref{sec:method}. The $E_{a,i}$'s for all
reactions except $\mathcal{R}_6$ -- $\mathcal{R}_9$ and $\mathcal{R}_{13}$ (due to 
zero nominal values for $E_a$)
were considered to be uncertain and uniformly distributed in the interval: 
$[0.99E_{a,i}^\ast, 1.01E_{a,i}^\ast]$. The nominal values, $E_{a,i}^\ast$
corresponding to the different reaction rates are provided in~\cite{Yetter:1991}.

\subsection{Computing the active subspace}

@@

In Figure~\ref{fig:hd} (left), we illustrate the resulting eigenvalue spectrum. The second eigenvalue is
found to be roughly two orders of magnitude smaller than the first, indicating that the active subspace
is 1-dimensional. This is further confirmed by the SSP plots in Figure~\ref{fig:hd} (right). 
%
\begin{figure}[htbp]
 \begin{center}
   \includegraphics[width=0.45\textwidth]{./Figures/eig_33D}
   \includegraphics[width=0.45\textwidth]{./Figures/ssp_33D}
\caption{Left: Eigenvalue spectrum of $\hat{\mat{C}}$. Right: SSP of the computed active subspace, and the
corresponding 1-D linear fit.} 
\label{fig:hd}
\end{center}
\end{figure}
%
Specifically
in the case of gradient-based approach, the SSP exhibits a linear variation that is reasonably
captured by its linear fit, $\tilde{G}$. Whereas, the SSP based on the gradient-free approach
is relatively more scattered. Consequently, the linear fit exhibits a discrepancy with the corresponding
fit for the gradient-based SSP.

@@

\subsection{GSA consistency check}

The normalized activity scores ($\tilde{\nu}_{i,r}$) based on the dominant eigenspace using the two
approaches (gradient-based and gradient-free) are compared with the total Sobol' indices
in Figure~\ref{fig:as_33D}. Note that the Sobol' indices were computed using the verified
1-dimensional surrogate in the active subspace for the gradient-based approach. 
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/as_33D_new}
\caption{A bar-graph illustrating individual activity scores for the uncertain $A_i$' (left) and $E_{a,i}$'s (right).}
\label{fig:as_33D}
\end{center}
\end{figure}
%
Several useful inferences can be drawn from the above plot. Firstly, the total Sobol indices are found to be
consistent
with the normalized activity scores evaluated using the gradient-based approach. Whereas, the gradient-free
approach yields consistent results for the $A_i$'s, it does not capture the sensitivity associated with
$E_{a,15}$. While this observation bolsters our confidence in the gradient-based approach, it clearly
underscores the shortcoming of the gradient-free approach. Numerical errors are incurred when approximating
the model gradients using the local linear approximation in this case. 
Secondly, our findings indicate that   
the ignition delay is predominantly sensitive towards $A_1$, $A_9$, $E_{a,1}$, and $E_{a,15}$ and 
moderately sensitive towards $A_{15}$ and $A_{17}$. Sensitivity towards the remaining uncertain inputs is
found to be low or negligible. 
%Note that this observation can also be exploited for reducing the 
%dimensionality of the problem from 33 inputs to 4-6 inputs depending upon the required level of accuracy.
%The sensitivity-driven dimension reduction for uncertainty quantification has been
%discussed in our earlier effort~\cite{Vohra:2018}. In this work, however, we focus on dimension reduction using the active 
%subspace approach which is shown to yield even greater scope for dimension reduction i.e. from 33 to 1. Furthermore, we have
%shown that the dominant eigenspace can be used to approximate global sensitivity measures with no additional effort.

### 2018-08-20 Paper/intro.tex
..

We seek to understand the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on the ignition delay. The $A_i$'s associated with all
reactions and the $E_{a,i}$'s with non-zero nominal estimates
are considered to be uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem computationally challenging due to the large number of 
uncertain parameters.  
To address this challenge, we focus on reducing the dimensionality
of the problem by computing the aforementioned active subspace.
This involves repeated evaluations of the gradient of a model output with
respect to the input parameters. Estimating the
gradients can however be computationally challenging in high-dimensional settings. 
Hence, in this work, we further explore the suitability of a gradient-free
approach that involves a regression-based local linear approximation of
the model output. \alennote{I don't understand what the next sentence is 
really trying to say here; I mean whatever we do involves approximating the
gradient, the question is how rough the approximation is, right?} 
It must be noted that since the gradient-free approach
relies on approximating the model gradients, it is expected to be less
robust than the gradient-based approach. The two approaches are based on
algorithms 1.1 and 1.2 in~\cite{Constantine:2015}. Both strategies
are implemented in an iterative manner i.e. a new set of model evaluations are
generated, and gradient evaluations are performed at each iteration until 
a reasonably accurate approximation to the active subspace is achieved, as
discussed later in Section~\ref{sec:method}.

An alternate approach to dimension reduction involves computing the global
sensitivity measures associated with uncertain inputs of a model. Depending
upon the estimates of the sensitivity measures, only the important inputs are
varied for the purpose of uncertainty quantification (UQ). Sobol' indices are
commonly used as global sensitivity measures~\cite{Sobol:2001}. However,
obtaining converged estimates of the Sobol' indices typically requires tens of
thousands of model runs and is therefore impractical in situations involving
expensive model simulations and/or high-dimensional input parameter spaces.
Recently, the derivative-based global sensitivity measures~(DGSMs) were
developed to approximate upper bound on the Sobol' indices with much fewer
computations~\cite{Sobol:2009, Lamboni:2013}. It was noted
in~\cite{Diaz:2016,Constantine:2017} that DGSMs can be approximated by
exploiting their links with active subspaces. This led to definition of the so
called \emph{activity scores}. In Section~\ref{sec:assa}, we build on these
ideas to provide a complete analysis of links between Sobol indices, DGSMs, and
activity scores for functions of independent random inputs whos distribution
law belongs to the class of Boltzmann probability measures. 
It is worth mentioning that computing global sensitivity measures provides 
important information about a model that go beyond dimension reduction. By 
identifying parameters with significant impact on model uncertainty global 
sensitivity analysis can guide model calibration and risk assessment.

The main contributions of this paper are as follows: 
\begin{itemize}
\item 
Active subspace
discovery in the high-dimensional H$_2$/O$_2$ kinetics problem involving 33
uncertain inputs: \alennote{Say some interesting conclusion of the analysis;
e.g., we discover a 1D subspace for a 33-dimensional problem using a modest computational 
budget ...} The presented
analysis can also guide practitioners in area of chemical kinetics on using the
method of active subspaces as an efficient means of uncertainty propagation in
chemical systems.  
\item Comprehensive numerical investigation of the gradient-based and gradient-free approach: 
we investigate the suitability of both approach
for computing active subspaces, surrogate modeling, and approximating global
sensitivity measures through a comprehensive set of numerical experiments. Our
results reveal insight into the merits of the methods approach as well as
their shortcomings.  
\item Analysis of the links between  
global sensitivity measures: 
By connecting the recent advances in theory of variance based and
derivative based global sensitivity analysis to active subspaces, we provide a complete analysis
of the links between total Sobol' indices, DGSMs, and activity scores for a broad
class of probability distributions. Our analysis is concluded by a result quantifying
approximation errors incurred due to fixing unimportant parameters, deemed so by 
computing their activity scores.

@@

\alennote{Okay, I reorganized the sections a bit (new sec 2, 3); please read the following and
make sure it reflects the contents of the sections correctly.}
This article is organized as follows. In section~\ref{sub:ac}, a brief
theoretical background on the active subspace methodology is provided. In
section~\ref{sub:gsa}, it is shown that the activity scores provide a
reasonable approximation to the DGSMs especially in a high-dimensional setting.
Additionally, a relationship between the three global sensitivity measures,
namely, the activity scores, DGSMs, and the total Sobol' indices is
established. In section~\ref{sec:method}, a systematic framework for computing
the active subspace using an iterative gradient-based approach is provided.
Numerical results based on the gradient-free approach are compared with those
obtained using the gradient-based approach.  The active subspace is computed
for the H$_2$/O$_2$ reaction kinetics problem whereby only the $A_i$'s are
considered as uncertain. In this case, the two approaches are observed to yield
consistent results. In section~\ref{sec:app}, the gradient-based approach is
further implemented to compute the active subspace for the high-dimensional
H$_2$/O$_2$ reaction kinetics problem. The resulting active subspace is found
to be 1-dimensional thereby indicating enormous potential for computational
savings in this application. The active subspace is also computed using the
gradient-free approach. It is observed that although the uncertainty in the
ignition delay is reasonably approximated, the gradient-free approach fails to
capture the global sensitivity of all the uncertain inputs. Hence, a trade-off
between efficiency and accuracy between the two approaches is observed in this
case.  Finally, a summary and discussion based on our findings is included in
section~\ref{sec:conc}.

>>>>

We seek to understand the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on the ignition delay. The $A_i$'s associated with all
reactions and the $E_{a,i}$'s with non-zero nominal estimates
are considered to be uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem computationally challenging due to the large number of 
uncertain parameters.  
To address this challenge, we focus on reducing the dimensionality
of the problem by computing the aforementioned active subspace.
This involves repeated evaluations of the gradient of a model output with
respect to the input parameters. Estimating the
gradients can however be computationally challenging in high-dimensional settings. 
Hence, in this work, we further explore the suitability of a gradient-free
approach that involves a regression-based local linear approximation of
the model output as discussed later in~\ref{sub:gradfree}. The two approaches
 are based on
algorithms 1.1 and 1.2 in~\cite{Constantine:2015}. Both strategies
are implemented in an iterative manner i.e. a new set of model evaluations are
generated, and gradient evaluations are performed at each iteration until 
a reasonably accurate approximation to the active subspace is achieved, as
discussed later in Section~\ref{sec:method}.

An alternate approach to dimension reduction involves computing the global
sensitivity measures associated with uncertain inputs of a model. Depending
upon the estimates of the sensitivity measures, only the important inputs are
varied for the purpose of uncertainty quantification (UQ). Sobol' indices are
commonly used as global sensitivity measures~\cite{Sobol:2001}. However,
obtaining converged estimates of the Sobol' indices typically requires tens of
thousands of model runs and is therefore impractical in situations involving
expensive model simulations and/or high-dimensional input parameter spaces.
Recently, the derivative-based global sensitivity measures~(DGSMs) were
developed to approximate upper bound on the Sobol' indices with much fewer
computations~\cite{Sobol:2009, Lamboni:2013}. It was noted
in~\cite{Diaz:2016,Constantine:2017} that DGSMs can be approximated by
exploiting their links with active subspaces. This led to definition of the so
called \emph{activity scores}. In Section~\ref{sub:gsa}, we build on these
ideas to provide a complete analysis of links between Sobol indices, DGSMs, and
activity scores for functions of independent random inputs whose distribution
law belongs to the class of Boltzmann probability measures. 
It is worth mentioning that computing global sensitivity measures provides 
important information about a model that go beyond dimension reduction. By 
identifying parameters with significant impact on model uncertainty global 
sensitivity analysis can guide model calibration and risk assessment.

The main contributions of this paper are as follows: 
\begin{itemize}
\item 
Active subspace
discovery in a high-dimensional H$_2$/O$_2$ kinetics problem involving 33
uncertain inputs. The implementation of the methodology presented in this work
resulted in a 1-dimensional active subspace, indicating immense potential
for computational savings. The presented
analysis can also guide practitioners in area of chemical kinetics on using the
method of active subspaces as an efficient means for uncertainty propagation.  
\item Comprehensive numerical investigation of the gradient-based and gradient-free approach: 
we investigate the suitability of both approach
for computing active subspaces, surrogate modeling, and approximating global
sensitivity measures through a comprehensive set of numerical experiments. Our
results reveal insight into the merits of the methods approach as well as
their shortcomings.  
\item Analysis of the links between  
global sensitivity measures: 
By connecting the recent advances in theory of variance based and
derivative based global sensitivity analysis to active subspaces, we provide a complete analysis
of the links between total Sobol' indices, DGSMs, and activity scores for a broad
class of probability distributions. Our analysis is concluded by a result quantifying
approximation errors incurred due to fixing unimportant parameters, deemed so by 
computing their activity scores.

@@

This article is organized as follows. In section~\ref{sub:ac}, a brief
theoretical background on the active subspace methodology is provided. In
section~\ref{sub:gsa}, it is shown that the activity scores provide a
reasonable approximation to the DGSMs especially in a high-dimensional setting.
Additionally, a relationship between the three global sensitivity measures,
namely, the activity scores, DGSMs, and the total Sobol' indices is
established. In section~\ref{sec:method}, a systematic framework for computing
the active subspace using an iterative gradient-based approach is provided.
Numerical results based on the gradient-free approach are compared with those
obtained using the gradient-based approach.  The active subspace is computed
for the H$_2$/O$_2$ reaction kinetics problem whereby only the $A_i$'s are
considered as uncertain. In this case, the two approaches are observed to yield
consistent results. In section~\ref{sec:app}, the gradient-based approach is
further implemented to compute the active subspace for the high-dimensional
H$_2$/O$_2$ reaction kinetics problem. The resulting active subspace is found
to be 1-dimensional thereby indicating enormous potential for computational
savings in this application. The active subspace is also computed using the
gradient-free approach. It is observed that although the uncertainty in the
ignition delay is reasonably approximated, the gradient-free approach fails to
capture the global sensitivity of all the uncertain inputs. Hence, a trade-off
between efficiency and accuracy between the two approaches is observed in this
case.  Finally, a summary and discussion based on our findings is included in
section~\ref{sec:conc}.
